\chapter{Predicate Abstraction and Concurrency}
\section{Introduction}
Predicate Abstraction is used to automatically verify: C device drivers (SLAM project at MSR), distributes concurrent algorithms, biological systems (PRISM at Oxford), security properties.
\section{Theory}
\subsection{Domain}
\subsubsection{Concreate Domain}
$Lab \to p(Store)$ \textbf{TODO: powerset symbol?}\\
e.g. $2 \to \{\{x=1,y=3\},\{x=2,y=2\}\}$
\subsubsection{Logical Domain}
Capture the set of stores with a FOL formula. Example:\\
$1 \to x=1 \wedge y=2\\
2 \to (x=1\wedge y=3) \vee (x=2\wedge y=2)$
Then a Galois Connection between $Lab \to p(Store)$ and $Lab \to FOL$ can be setup.
\subsection{Concrete Transformers}
$F^{FOL}:(Lab \to FOL) \to (Lab \to FOL)$\\
$F^{FOL}(m)l = \begin{cases} true &\mbox{if l is initial label}\\ 
\bigvee_{(l',action,l)}\llbracket action \rrbracket_{FOL}(m(l')) & \mbox{otherwise}\end{cases}$
\subsubsection{Effect of action on a FOL formula}
$\llbracket action \rrbracket_{FOL}: FOL \to FOL$\\Examples: \\
$\llbracket skip \rrbracket_{FOL}(x=1 \wedge y=2) = (x=1 \wedge y=2)\\ \llbracket (x > 5)\rrbracket_{FOL}(x>0 \wedge y=2)=(x>5 \wedge y=2)\\ \llbracket (x:=7) \rrbracket_{FOL}((x<0) \wedge (x=5)) = (x=7)$
\subsubsection{Strongest post condition}
The \underline{best transformer} for $\llbracket action \rrbracket_{FOL}$ is the strongest post-condition (sp).\\
$\llbracket action \rrbracket_{FOL}(\psi) = sp(\psi, action)$\\
With $sp(\psi, action) = \psi'$ The formula $\psi'$ si the most precise (strongest) set of successor stores of $\psi$ as determined by action.\\
$sp(\psi, skip) = \psi\\
sp(\psi, b) = \psi \wedge b\\
sp(\psi, x:=a) = \exists v:x=a[v/x] \wedge \psi[v/x]$\\
Quantifiers are difficult to handle for automated reasoning engines and many decidable fragments do not allow quantifiers.
\subsubsection{Weakest Pre-Condition}
$wp(\psi, skip) = \psi\\ wp(\psi,b) = b \Rightarrow \psi\\ wp(\psi, x:=a) = \psi[a/x]$\\
\textbf{No quantifiers!}
\subsubsection{linking sp with wp}
\begin{theorem}
$sp(\psi, action) \Rightarrow \psi' \equiv \psi \Rightarrow wp(\psi', action)$
\end{theorem}
Using weakest preconditions we get rid of quantifier. \textbf{TODO}: check again. 
\subsubsection{Undecidability}
FOL is undecidable. If $\psi$ in FOL is not valid, the no algorithm is guaranteed to terminate. Can't guarantee termination of iterating $F^{FOL}$.
\section{Predicate Abstraction recipe}
\begin{enumerate}
\item Come up with an \underline{abstract domain} (based on the type of properties you want to prove)
\item Define abstract semantics \underline{for the programming language} w.r.t. the abstract domain from step 1.
\begin{itemize}
\item define the abstract transformers (effect of statement on the domain)
\item prove that the abstract semantics are \underline{sound} w.r.t \underline{concrete semantics} of the programming language
\end{itemize}
\item iterate abstract transformers over the abstract domain until fixed point
\end{enumerate}
\subsection{Abstract domain}
We define $F$ to be a \underline{finite} set of predicates $F=\{\phi_1, \phi_2,... \phi_n\}$ where the predicates are selected from a \underline{decidable} logical fragment. Examples for the predicates are: \\$\phi_1 := x > 0 \wedge x = (x+1)\\ \phi_2:=z+y>5 \wedge x=(y+y)$ (over theory of integers).\\
Using the predicates we define a \underline{cube} which is a \underline{conjuction} of predicates and negated predicates where each $\phi_i$ appears \underline{at most once}. Cubes can be represented as vectors of size n where the value at index $i$ indicates if $\phi_i$ is negated (0), positive (1) or missing (T).\\ \\
Our resulting abstract domain is $Lab \to Cube_F$ where $Cube_F$ is the set of all possible cubes formed over the predicates in $F$.\\ \\ 
Let's define $\sqsubseteq_{pa}, \sqcap_{pa}, \sqcup_{pa}$:
\begin{itemize}
    \item Ordering $\sqsubseteq_{pa}$: 
        \begin{itemize}
            \item $a \sqsubseteq_{pa} b$ compares the elements pointwise
            \item $a \sqsubseteq_{pa} b $ means $a \Rightarrow b$, but not vice versa
            \item Examples:\\
                $101 \sqsubseteq_{pa} 100 = no\\ 100 \sqsubseteq_{pa} 101 = no\\T01 \sqsubseteq_{pa} 10T = no\\011 \sqsubseteq_{pa} 01T = yes\\01T \sqsubseteq_{pa} 011 = no$
        \end{itemize}
    \item Greates lower bound $\sqcap_{pa}$
        \begin{itemize}
            \item keep $\phi_k$ unsell $\phi_k$ appears in a, and $\neg \phi_k$ appears in b (or vice versa) in which case $a \sqcap_{pa} b$ returns false
            \item Examples: $101 \sqcap_{pa} 100 = false\\101 \sqcap_{pa} 10T = 101\\ T01 \sqcap_{pa} 10T = 101\\ 00T \sqcap_{pa} 01T = false$
        \end{itemize}
    \item Least upper bound $\sqcup_{pa}$
        \begin{itemize}
            \item predicate (or negation) is kept if it appears in both cubes. 
            \item if all predicates are ommited then $\sqcup_{pa}$ returns true
            \item $a \vee b \Rightarrow a \sqcup_{pa} b$
            \item Examples:\\ $101 \sqcup_{pa} 100 = 10T\\101 \sqcup_{pa} 010 = true\\T01 \sqcup_{pa} 10T = T0T\\true \sqcup_{pa} 011 = true$
        \end{itemize}
\end{itemize}
\subsubsection{Connection Domains - Galois Connections}
Next we setup a Galois connection between FOL and $Cube_F$
Concretization: $\gamma: Cube_F \to FOL\\ \gamma(cube) = cube$\\ $\alpha$ is defined by $\gamma$ and vice versa. So: \\
$\alpha(\psi) = \sqcap_{pa}\{cube | \psi \sqsubseteq_{FOL}\gamma(cube)\}$ \textbf{TODO:} where is this definition from?\\
In FOL, $\sqsubseteq_{FOL}$ is: $\alpha(\psi) = \sqcap_{pa}\{cube | \psi \Rightarrow \gamma(cube)\}$ By substitution of $\gamma$ we get:\\
\underline{$a(\psi)=\sqcap_{pa}\{cube|\psi \Rightarrow cube\}$}\\
Example:\\
$F=\{(2x-y), (y*y>4), (x+y<5)\}\\
\alpha(x=1 \wedge y=3) = \sqcap_{pa}\{0TT, T1T, TT1, 01T, 0T1, T11, 011\} = 011 = \neg(2x-y>0)\wedge (y*y>4)\wedge x+y<5$\\
\textbf{Optimization}: instead of $\alpha(\psi)=\sqcap_{pa}\{cube | \psi \Rightarrow cube\}$ use $\alpha(\psi)=\sqcap_{pa}\{literal|\psi \Rightarrow literal\}$
$literal \iff cube$ has only one value different that T in the correpsonding array. Example: 0TT, T1T. \textbf{TODO: no idea here, page 28}\\
\textbf{TODO: why is this a galois connection?}
\subsection{Abstract Transformers}
\textbf{TODO: pretty cool construction of abstract transformer}\\
$\llbracket action \rrbracket_{FOL}(\psi) = sp(\psi, action)$ is the best concrete transformer. \textbf{Why?} Now lets defined the \underline{best abstract transformer}: $\llbracket action \rrbracket_{pa}:Cube_F \to Cube_F$
\begin{align}
    \llbracket action \rrbracket_{pa}(c) 
    &= \alpha \circ \llbracket action \rrbracket_{FOL} \circ \gamma(c)
    \\&= \alpha \circ \llbracket action \rrbracket_{FOL}(c) && \text{substitution of $\gamma$}
    \\&= \sqcap_{pa}\{literal | \llbracket action \rrbracket_{FOL}(c) \Rightarrow literal\} && \text{definition of $\alpha$}
    \\&=\sqcap_{pa}\{literal | sp(c,action) \Rightarrow literal\} &&\text{definition of $\llbracket action \rrbracket_{pa}$}
    \\&=\sqcap_{pa}\{literal | c \Rightarrow wp(literal, action)\} &&\text{by sp to wp connection}
\end{align}
So to compute the tranformer, we need to check whether the formula $c \Rightarrow wp(literal, action)$ is \underline{valid}. This is done with an SMT solver and it's therefore desirable that the formulas are in some decidable logical fragment.
\subsubsection{Key points}
\begin{itemize}
    \item for predicate abstraction the transformers are proved correct once and for all
    \item then we can instantiate predicate abstraction with any predicates and don't need to prove correctness
    \item in that sense predicate abstraction is a parametric framework
    \item the challenge is to find the sufficient predicates
\end{itemize}
\subsection{Iterate to a fixed point}
see page 33.
\subsection{Summary}
\begin{itemize}
    \item we used wp for the best transformer because it introduces no quantifiers
    \item Best abstract transformer is sound by construction
    \item an important challenge is to find the sufficient predicates F to verify the program
\end{itemize}
\subsubsection{An alternative to fidex-point iteration}
\textbf{TODO}: important / good idea?\\ 
Input: a program $P$, a set of predicates $F$ and a property $S$ to verify. 
\begin{enumerate}
    \item Build a boolean program $B(P,F)$\\
        Program contains only boolean variables, one for each predicate in F.
    \item Check that $B(P,F)$ verifies the property $S$. \\
        If the property holds for $B(P,F)$ then it also holds for P.
\end{enumerate}
\textbf{TODO: pages 36-37}
\section{Predicate Abstraction for Modern Concurrency}
\textbf{TODO: } check again. especially the "boolean program" part. 
cube is $O(3^{length})$ used extrapolition to work on cubes $\ll cube$ 
